Fluent statistical computing interfaces for biological data analysis

### 2020-10-28 Rmd/01-intro.Rmd
capitalisation

Exploratory data analysis (EDA) is a vital element of the modern 
statistical workflow - it is an analyst’s first pass at 
understanding their data; revealing all its messes and 
uncovering hidden insights [@Tukey1977-pn; @Grolemund2017-uc]. 
It is an iterative process involving computation and visualization,
leading to new hypotheses that can be tested and formalised using 
statistical modelling (figure \@ref(fig:wflow-cycle)). As datasets grow in complexity and become
increasingly heterogeneous and multidimensional, 
the use of EDA becomes vital to ensure the integrity and quality of analysis 
outputs. This is certainly true in high-throughput biological 
data science, where constraints on computation time and memory, in addition
to the analyst's time, makes EDA difficult and neglected, which impacts the robustness and reliability of any downstream analysis.

@@

This thesis focuses on core aspects of EDA as part of a biological data
science workflow: wrangling, integration and visualisation, with a focus 
on applications to genomics and transcriptomics. To begin we discuss wrangling
biological data using a coherent representation and programming interface 
(figure \@ref(fig:wflow-wrangle)). Section \@ref(sec:tidygranges)
introduces a grammar-based framework for transforming genomics data that is 
described in chapter \@ref(ch:plyranges). We then look at integrating data and 
model outputs over genomic regions to gain biological insight (\@ref(fig:wflow-integrate)). Section \@ref(sec:integrate) introduces a 
framework for incorporating genomic regions  over multiple assays, described in 
chapter \@ref(ch:fluentGenomics), while  \@ref(sec:coverage) discusses finding 'interesting' genomic regions via combining multiple summaries of a single assay, described in \@ref(ch:intron). Next, we 
consider the challenges in  visualising  high dimensional data (figure \@ref(fig:wflow-explore)). Section  \@ref(sec:va) introduces an interactive 
visualisation approach for  understanding non-linear dimension reduction 
techniques described in chapter  \@ref(ch:tsne). Lastly, in 
chapter \@ref(ch:conclusion) describes the outputs of the thesis and plans for future developments.

@@

It is unclear whether the *fluent* interfaces as implemented using the
*tidy data* framework can be more generally applied and useful
in fields such as high-throughput biology where domain specific semantics are 
required (figure \@ref(fig:wflow-wrangle)). This is particularly true in the Bioconductor ecosystem, where much 
thought has gone into the design of data structures that enable interoperability
between different tools, biological assays and analysis goals [@Huber2015-ei].

@@

It is rare that a biological data analysis will involve
a single measurement assay or that only one aspect of a measurement assay
will be of interest to the biological question under study (figure \@ref(fig:wflow-integrate)). While there are 
many approaches to integrating data sets from multiple assays using multivariate
statistical techniques [@Meng2016-pz; @Stein-OBrien2018-bk] and data structures to represent them [@Ramos2017-kn], there has been little thought given to the interoperability between these approaches and the **tidyverse**. In chapter \@ref(ch:fluentGenomics)
we describe a simple end-to-end workflow for integrating 
results along the genome using **plyranges**. This workflow shows that
our grammar based approach does not impair interoperability between
the **tidyverse** and Bioconductor approaches, and in fact they work seamlessly
together. This chapter has been published as @Lee2020-ie.

@@

In chapter \@ref(ch:intron) we explore the limits of the *tidy data*
semantic by extending **plyranges** to analyse coverage estimated on RNA-seq
data by developing a new software tool called **superintronic**. We show that the
long-form tidy representation is an effective way of combining the experimental
design and reference annotations into a single genomic data structure for
exploration. We use **superintronic** to develop a framework for discovering
interesting regions of coverage and apply our approach to integrating intron
signal from RNA-seq data. This chapter is based on my software and analysis
contributions to the preprint @Lee2019-mf.

@@

In the final chapter, we move away from data wrangling and towards the 
integration of visualisation with model-based summaries of high-dimensional
data sets (figure \@ref(fig:wflow-explore)). We focus on a common tool for EDA 
(especially applied to single-cell transcriptomics): non-linear dimension 
reduction (NLDR). We consider the incorporation of interactive and dynamic 
graphics to assist analysts in using  NLDR techniques for cluster orientation 
tasks. In particular, we advocate  for the use of tours [@Cook1995-bi] alongside
an NLDR visualisation to  highlight potential pitfalls  and distortions obtained
from an NLDR method. This approach acknowledges that there is no 'one' best 
visualisation or dimension reduction for a high-dimensional dataset, and we 
often want to  have an understanding of both the global and local structure 
within our data.

>>>>

Exploratory data analysis (EDA) is a vital element of the modern 
statistical workflow - it is an analyst’s first pass at 
understanding their data; revealing all its messes and 
uncovering hidden insights [@Tukey1977-pn; @Grolemund2017-uc]. 
It is an iterative process involving computation and visualization,
leading to new hypotheses that can be tested and formalised using 
statistical modelling (Figure \@ref(fig:wflow-cycle)). As datasets grow in complexity and become
increasingly heterogeneous and multidimensional, 
the use of EDA becomes vital to ensure the integrity and quality of analysis 
outputs. This is certainly true in high-throughput biological 
data science, where constraints on computation time and memory, in addition
to the analyst's time, makes EDA difficult and neglected, which impacts the robustness and reliability of any downstream analysis.

@@

This thesis focuses on core aspects of EDA as part of a biological data
science workflow: wrangling, integration and visualisation, with a focus 
on applications to genomics and transcriptomics. To begin we discuss wrangling
biological data using a coherent representation and programming interface 
(Figure \@ref(fig:wflow-wrangle)). Section \@ref(sec:tidygranges)
introduces a grammar-based framework for transforming genomics data that is 
described in Chapter \@ref(ch:plyranges). We then look at integrating data and 
model outputs over genomic regions to gain biological insight (\@ref(fig:wflow-integrate)). Section \@ref(sec:integrate) introduces a 
framework for incorporating genomic regions  over multiple assays, described in 
Chapter \@ref(ch:fluentGenomics), while  \@ref(sec:coverage) discusses finding 'interesting' genomic regions via combining multiple summaries of a single assay, described in \@ref(ch:intron). Next, we 
consider the challenges in  visualising  high dimensional data (Figure \@ref(fig:wflow-explore)). Section  \@ref(sec:va) introduces an interactive 
visualisation approach for  understanding non-linear dimension reduction 
techniques described in Chapter  \@ref(ch:tsne). Lastly, in 
Chapter \@ref(ch:conclusion) describes the outputs of the thesis and plans for future developments.

@@

It is unclear whether the *fluent* interfaces as implemented using the
*tidy data* framework can be more generally applied and useful
in fields such as high-throughput biology where domain specific semantics are 
required (Figure \@ref(fig:wflow-wrangle)). This is particularly true in the Bioconductor ecosystem, where much 
thought has gone into the design of data structures that enable interoperability
between different tools, biological assays and analysis goals [@Huber2015-ei].

@@

It is rare that a biological data analysis will involve
a single measurement assay or that only one aspect of a measurement assay
will be of interest to the biological question under study (Figure \@ref(fig:wflow-integrate)). While there are 
many approaches to integrating data sets from multiple assays using multivariate
statistical techniques [@Meng2016-pz; @Stein-OBrien2018-bk] and data structures to represent them [@Ramos2017-kn], there has been little thought given to the interoperability between these approaches and the **tidyverse**. In Chapter \@ref(ch:fluentGenomics)
we describe a simple end-to-end workflow for integrating 
results along the genome using **plyranges**. This workflow shows that
our grammar based approach does not impair interoperability between
the **tidyverse** and Bioconductor approaches, and in fact they work seamlessly
together. This chapter has been published as @Lee2020-ie.

@@

In Chapter \@ref(ch:intron) we explore the limits of the *tidy data*
semantic by extending **plyranges** to analyse coverage estimated on RNA-seq
data by developing a new software tool called **superintronic**. We show that the
long-form tidy representation is an effective way of combining the experimental
design and reference annotations into a single genomic data structure for
exploration. We use **superintronic** to develop a framework for discovering
interesting regions of coverage and apply our approach to integrating intron
signal from RNA-seq data. This chapter is based on my software and analysis
contributions to the  @Lee2019-mf.

@@

Finally, we move away from data wrangling and towards the 
integration of visualisation with model-based summaries of high-dimensional
data sets (Figure \@ref(fig:wflow-explore)). We focus on a common tool for EDA 
(especially applied to single-cell transcriptomics): non-linear dimension 
reduction (NLDR). We consider the incorporation of interactive and dynamic 
graphics to assist analysts in using  NLDR techniques for cluster orientation 
tasks. In particular, we advocate  for the use of tours [@Cook1995-bi] alongside
an NLDR visualisation to  highlight potential pitfalls  and distortions obtained
from an NLDR method. This approach acknowledges that there is no 'one' best 
visualisation or dimension reduction for a high-dimensional dataset, and we 
often want to  have an understanding of both the global and local structure 
within our data.

### 2020-10-28 Rmd/02-plyranges.Rmd
capitalisation

The **plyranges** DSL is built on the core Bioconductor data structure
_GRanges_, which is a constrained table, with fixed columns
for the chromosome, start and end coordinates, and the strand, along
with an arbitrary set of additional columns, consisting of
measurements or metadata specific to the data type or experiment
(figure \@ref(fig:GRanges)).  _GRanges_ balances flexibility with formal
constraints, so that it is applicable to virtually any genomic
workflow, while also being semantically rich enough to support
high-level operations on genomic ranges. As a core data structure,
_GRanges_ enables interoperability between **plyranges** and the rest of
Bioconductor. Adhering to a single data structure simplifies the API
and makes it easier to learn and understand, in part because
operations become endomorphic, i.e., they return the same type as
their input.

@@

To support merging, our algebra recasts finding overlaps or 
nearest neighbors between two
genomic regions as variants of the relational join operator. A join
acts on two _GRanges_ objects: _x_ and _y_. The join operator
is relational in the sense that metadata from the _x_ and _y_
ranges are retained in the joined range.  All join operators in the
**plyranges** DSL generate a set of hits based on overlap or proximity
of ranges and use those hits to merge the two datasets in different
ways. There are four supported matching algorithms: _overlap_,
_nearest_, _precede_, and _follow_ (figure 
\@ref(fig:olaps-fig)). We can further restrict the
matching by whether the query is completely _within_ the subject, and
adding the _directed_ suffix ensures that matching ranges have the
same direction (strand).

@@

Since the _GRanges_ object is a tabular data structure, our grammar includes
operators to filter, sort and aggregate by columns in a _GRanges_. These 
operations can be performed over partitions formed
using the `group_by()` modifier. Together with our algebra for arithmetic and
merging, these operations conform to the semantics and syntax of the 
**dplyr** grammar. Consequently, **plyranges** code is generally more compact
than the equivalent **GenomicRanges** code (figure \@ref(fig:code-comparison)).

@@

Finally, the overlap inner join is used to restrict the chromosome 10
coverage islands, to the islands that are contained in the 5000nt region that surrounds the max peak (figure \@ref(fig:peak-viz)).

@@

We then compute the transition-transversion ratio over each chromosome using
`group_by()` in combination with `summarize()` (figure \@ref(fig:titv-viz)).

>>>>

The **plyranges** DSL is built on the core Bioconductor data structure
_GRanges_, which is a constrained table, with fixed columns
for the chromosome, start and end coordinates, and the strand, along
with an arbitrary set of additional columns, consisting of
measurements or metadata specific to the data type or experiment
(Figure \@ref(fig:GRanges)).  _GRanges_ balances flexibility with formal
constraints, so that it is applicable to virtually any genomic
workflow, while also being semantically rich enough to support
high-level operations on genomic ranges. As a core data structure,
_GRanges_ enables interoperability between **plyranges** and the rest of
Bioconductor. Adhering to a single data structure simplifies the API
and makes it easier to learn and understand, in part because
operations become endomorphic, i.e., they return the same type as
their input.

@@

To support merging, our algebra recasts finding overlaps or 
nearest neighbors between two
genomic regions as variants of the relational join operator. A join
acts on two _GRanges_ objects: _x_ and _y_. The join operator
is relational in the sense that metadata from the _x_ and _y_
ranges are retained in the joined range.  All join operators in the
**plyranges** DSL generate a set of hits based on overlap or proximity
of ranges and use those hits to merge the two datasets in different
ways. There are four supported matching algorithms: _overlap_,
_nearest_, _precede_, and _follow_ (Figure 
\@ref(fig:olaps-fig)). We can further restrict the
matching by whether the query is completely _within_ the subject, and
adding the _directed_ suffix ensures that matching ranges have the
same direction (strand).

@@

Since the _GRanges_ object is a tabular data structure, our grammar includes
operators to filter, sort and aggregate by columns in a _GRanges_. These 
operations can be performed over partitions formed
using the `group_by()` modifier. Together with our algebra for arithmetic and
merging, these operations conform to the semantics and syntax of the 
**dplyr** grammar. Consequently, **plyranges** code is generally more compact
than the equivalent **GenomicRanges** code (Figure \@ref(fig:code-comparison)).

@@

Finally, the overlap inner join is used to restrict the chromosome 10
coverage islands, to the islands that are contained in the 5000nt region that surrounds the max peak (Figure \@ref(fig:peak-viz)).

@@

We then compute the transition-transversion ratio over each chromosome using
`group_by()` in combination with `summarize()` (Figure \@ref(fig:titv-viz)).

### 2020-10-28 Rmd/04-intron.Rmd
finishing touches
capitalisation

```{r cov-trace, echo = FALSE, fig.cap="(ref:example-trace)}
knitr::include_graphics(here::here("img", "peak-viz-1.png"))
```

@@

**superintronic** is an R package used for estimating, representing,
and visualising per-base coverage scores, that can then be 
flexibly summarised over factors within an experimental design and collapsed
over regions of the genome using our companion package **plyranges**. The
aspects of this combined workflow are summarised in figure \@ref(fig:si-fig).

@@

The gene annotation files were
obtained as GFF files from RefSeq and used to construct the exonic
and intronic parts of each gene as _GRanges_ object using **superintronic**. 
We further filtered genes that had a single exon or genes that were overlapping
others in the annotation and that were not on the main contigs of the reference genome (i.e. excluding mitochondria) to simplify our analysis and reduce any coverage ambiguity. This left 18,270 genes available for 
computing data descriptors on in order to detecting IR-like coverage traces. Across each combination of genotype and line, we first $\log_2$ transformed the coverage score with an offset of one, and then intersected the coverage _GRanges_ with exonic and intronic features of each gene. For each gene,
the mean and standard deviation of the log-transformed coverage score weighted by the number of bases covered were computed over all intron and exon parts within the groupings of 
genotype and line. We also computed the bases above descriptor for intron
features. It refers to the total number of bases within an intron that has a score above the overall average exon coverage score. The scatter plot 
matrix view of these descriptors is shown in figure \@ref(fig:si-splom)
for a single biological group in the experiment. Using these views and
by summarising over the data descriptors, we came up with thresholds
for finding genes that have IR-like traces within each biological group.

@@

We selected genes with the following thresholds:  genes have an average exon log-coverage greater than the mean of average exon log-coverage values across all genes, have an average intron log-coverage greater than the average standard deviation of exon log-coverage values over all genes, and
the standard deviation of intron log-coverage values is twice the standard
deviation of exon log-coverage values. That is, we are selecting genes that
are expressed but  have large average intron expression that is more variable than the gene's exon expression. This results in a total of 86 genes selected
to link back to their underlying coverage traces, with the overlaps shown
in the UpSet plot in figure \@ref(fig:si-upset) [@Lex2014-bh]. The procedure produces gene coverage traces with known minor class splice sites affected by the knockout procedure such as ccdc43 and nat15 (figures \@ref(fig:si-cvg-ccdc43) and \@ref(fig:si-cvg-nat15)), as well as some that appear to have U12 intron retention like events such as mapk3 or tspan31 (figures \@ref(fig:si-cvg-mapk3) and \@ref(fig:si-cvg-tspan31)) that affect other parts of the gene.

@@

Our zebrafish workflow shows that our approach using **superintronic** 
and **plyranges** is able to uncover interesting biological signals in a
purely data-driven manner. We did not include additional information
that could have been useful when deriving our selected genes, such as
sequence motifs for U12 class of introns, or exploit the experimental
design to find differential IR like profiles. However, if that was of interest,
one could look at the overlaps, like we did in figure \@ref(fig:si-upset),
or combine our data descriptors with external estimates using **limma** 
[@Ritchie2015-mj], like our proposed index method in @Lee2019-mf. The gene
candidates obtained by our thresholds have been validated by the Heath lab
using qPCR.

>>>>

```{r cov-trace, echo = FALSE, fig.cap="(ref:example-trace)"}
knitr::include_graphics(here::here("img", "peak-viz-1.png"))
```

@@

**superintronic** is an R package used for estimating, representing,
and visualising per-base coverage scores, that can then be 
flexibly summarised over factors within an experimental design and collapsed
over regions of the genome using our companion package **plyranges**. The
aspects of this combined workflow are summarised in Figure \@ref(fig:si-fig).

@@

The gene annotation files were
obtained as GFF files from RefSeq and used to construct the exonic
and intronic parts of each gene as _GRanges_ object using **superintronic**. 
We further filtered genes that had a single exon or genes that were overlapping
others in the annotation and that were not on the main contigs of the reference genome (i.e. excluding mitochondria) to simplify our analysis and reduce any coverage ambiguity. This left 18,270 genes available for 
computing data descriptors on in order to detecting IR-like coverage traces. Across each combination of genotype and line, we first $\log_2$ transformed the coverage score with an offset of one, and then intersected the coverage _GRanges_ with exonic and intronic features of each gene. For each gene,
the mean and standard deviation of the log-transformed coverage score weighted by the number of bases covered were computed over all intron and exon parts within the groupings of 
genotype and line. We also computed the bases above descriptor for intron
features. It refers to the total number of bases within an intron that has a score above the overall average exon coverage score. The scatter plot 
matrix view of these descriptors is shown in Figure \@ref(fig:si-splom)
for a single biological group in the experiment. Using these views and
by summarising over the data descriptors, we came up with thresholds
for finding genes that have IR-like traces within each biological group.

@@

We selected genes with the following thresholds:  genes have an average exon log-coverage greater than the mean of average exon log-coverage values across all genes, have an average intron log-coverage greater than the average standard deviation of exon log-coverage values over all genes, and
the standard deviation of intron log-coverage values is twice the standard
deviation of exon log-coverage values. That is, we are selecting genes that
are expressed but  have large average intron expression that is more variable than the gene's exon expression. This results in a total of 86 genes selected
to link back to their underlying coverage traces, with the overlaps shown
in the UpSet plot in Figure \@ref(fig:si-upset) [@Lex2014-bh]. The procedure produces gene coverage traces with known minor class splice sites affected by the knockout procedure such as ccdc43 and nat15 (figures \@ref(fig:si-cvg-ccdc43) and \@ref(fig:si-cvg-nat15)), as well as some that appear to have U12 intron retention like events such as mapk3 or tspan31 (figures \@ref(fig:si-cvg-mapk3) and \@ref(fig:si-cvg-tspan31)) that affect other parts of the gene.

@@

Our zebrafish workflow shows that our approach using **superintronic** 
and **plyranges** is able to uncover interesting biological signals in a
purely data-driven manner. We did not include additional information
that could have been useful when deriving our selected genes, such as
sequence motifs for U12 class of introns, or exploit the experimental
design to find differential IR like profiles. However, if that was of interest,
one could look at the overlaps, like we did in Figure \@ref(fig:si-upset),
or combine our data descriptors with external estimates using **limma** 
[@Ritchie2015-mj], like our proposed index method in @Lee2019-mf. The gene
candidates obtained by our thresholds have been validated by the Heath lab
using qPCR.

### 2020-10-28 Rmd/05-embeddings.Rmd
capitalisation

From the video linked in figure \@ref(fig:gaussian), we learn that t-SNE has 
correctly split out each cluster and laid them out in a star like formation. 
This is agrees with the tour view,
where once we start the animation, the five clusters begin to appear but
generally points are more concentrated in the projection view compared to 
the t-SNE layout (figure \@ref(fig:gaussian)a). 
This can be seen via brushing the t-SNE view (figure \@ref(fig:gaussian)b).

@@

Next we view Gaussian clusters from the 
_Multi Challenge Dataset_, a benchmark simulation data set for clustering tasks
[@Rauber2009-vh]. This dataset has two Gaussian clusters with equal covariance 
embedded in 10-$d$, and a third cluster
with hierarchical structure. This cluster has two 3-$d$ clusters embedded in 
10-$d$, where the second cluster is subdivided into three smaller clusters, 
that are each equidistant from each other and have the same covariance structure.
From the video linked in figure \@ref(fig:hierarchical), we see 
that t-SNE has correctly identified the sub-clusters. However, their relative 
locations to each other is distorted, with the orange and blue groups being
far from each other in the tour view (figure \@ref(fig:hierarchical)a). We see 
in this case that is difficult
to see the sub-clusters in the tour view, however, once we zoom and highlight
they become more apparent (figure \@ref(fig:hierarchical)b). When we brush 
the sub-clusters in the t-SNE, their
relative placement is again exaggerated, with the tour showing that they
are indeed much closer than the impression the t-SNE view gives.

@@

Next we explore some simulated noisy tree structured data (figure
\@ref(fig:fake-trees)). Our interest here is how t-SNE visualisations break
topology of the data, and then seeing if we can resolve this by tweaking the
default parameters with reference to the global view of the data set. 
This simulation aims to mimic branching trajectories of cell differentiation:
if there were only mature cells, we would just see the tips of the branches 
which have a hierarchical pattern of clustering.

@@

Moreover, we run t-SNE using the default arguments on the complete data (this
keeps the first 50 PCs, sets the perplexity to equal 30 and performs random
initialisation). We then create a linked
tour with t-SNE layout with **liminal** as shown in figure \@ref(fig:trees-01).

@@

From the linked video, we see that the t-SNE view has been unable to 
recapitulate the topology of the tree - the backbone (blue) branch has been 
split into three fragments (figure \@ref(fig:trees-01)a). We can see 
this immediately via the linked highlighting over both
plots. If we click on the legend for the zero branch, the blue coloured points
on each view are highlighted and the remaining points are made transparent.
From, here it becomes apparent from the tour view that the blue branch forms
the backbone of the tree and is connected to all other branches. From the video 
it is easy to see that cluster sizes formed via t-SNE can be misleading; from the tour
view there is a lot of noise along the branches, while this does not appear to
be the case for the t-SNE result (figure \@ref(fig:trees-01)b).

From the first view, we modify the inputs to the t-SNE view, to try and produce
a better trade-off between local structure and retain the topology of the data.
We keep every parameter the same except that we initialise $Y$ with the first
two PCs (scaled to have standard deviation 1e-4) instead of the default random
initialisation and increase the perplexity from 30 to 100. We then combine
these results with our tour view as displayed in the linked video in the caption
of figure \@ref(fig:trees-02).

@@

The video linked in figure \@ref(fig:trees-02) shows that this selection of 
parameters results in the 
tips of the branches (the three black dots in figure \@ref(fig:fake-trees)a) 
being split into three clusters representing the terminal branches of the tree. 
However, there are perceptual issues following the placement of the three 
groupings on the t-SNE view that become apparent via simple linked brushing. 
If we brush the
tips of the yellow and brown branches (which appear to be close to each other
on the t-SNE view), we immediately see the placement is distorted in the t-SNE
view, and in the tour view these tips are at opposite ends of the tree 
(figure \@ref(fig:trees-02)b).
Although, this is a known issue of the t-SNE algorithm, we can easily identify
it via simple interactivity without knowing the inner workings of the method.

@@

Due to latency of the **liminal** interface we do a weighted sample of  the rows based on
cluster membership, leaving us with approximately 10 per cent of the original
data size - 4,590 cells. Although this is not ideal, it still allows us to
get a sense of the shape of the clusters as seen from the linked video in figure \@ref(fig:mouse-01). If one was interested in performing more in-depth cluster analysis we recommend an iterative approach of removing large clusters and then re-running the **liminal** view as a way finding more granular cluster structure.

@@

From the video linked in figure \@ref(fig:mouse-01), we learn that the embedding has mostly captured the
clusters relative location to each other to their location in high dimensional space, with a notable exception of points in cluster 3 and 10 as shown with
linked brushing (figure \@ref(fig:mouse-01)a). As expected, t-SNE mitigates the crowding problem that is
an issue for tour in this case, where points in clusters 2,4,6, and 11 are
clumped together in tour view, but are blown up in the embedding view (figure \@ref(fig:mouse-01)b). 
The tour appears to form a tetrahedron like shape, with points lying on
surface and along the vertices of the tetrahedron in 5-$d$ PCA space -
a phenomena that has also been observed in @Korem2015-af 
(figure \@ref(fig:mouse-01)c). Brushing on
the tour view, reveals points in cluster 9 that are diffuse in the embedding 
view, points in cluster 9 are relatively far away and spread apart from other 
clusters in the tour view, but has points placed in cluster 3 and 9 in the embedding
(figure \@ref(fig:mouse-01)d).

@@

From the video linked in figure \@ref(fig:mouse-02), we see that the 
expression of the marker genes, appear to separate
the previously clumped clusters 2,4,6, and 11 from the other clusters, 
indicating that these genes are expressed in all four clusters (figure \@ref(fig:mouse-02)a). After zooming, we can see a trajectory forming along
the clusters, while the axis view shows that magnitude of expression in 
the marker genes is similar across these separated clusters which is 
consistent with the results of marker gene analysis (figure \@ref(fig:mouse-02)b).

>>>>

From the video linked in Figure \@ref(fig:gaussian), we learn that t-SNE has 
correctly split out each cluster and laid them out in a star like formation. 
This is agrees with the tour view,
where once we start the animation, the five clusters begin to appear but
generally points are more concentrated in the projection view compared to 
the t-SNE layout (Figure \@ref(fig:gaussian)a). 
This can be seen via brushing the t-SNE view (Figure \@ref(fig:gaussian)b).

@@

Next we view Gaussian clusters from the 
_Multi Challenge Dataset_, a benchmark simulation data set for clustering tasks
[@Rauber2009-vh]. This dataset has two Gaussian clusters with equal covariance 
embedded in 10-$d$, and a third cluster
with hierarchical structure. This cluster has two 3-$d$ clusters embedded in 
10-$d$, where the second cluster is subdivided into three smaller clusters, 
that are each equidistant from each other and have the same covariance structure.
From the video linked in Figure \@ref(fig:hierarchical), we see 
that t-SNE has correctly identified the sub-clusters. However, their relative 
locations to each other is distorted, with the orange and blue groups being
far from each other in the tour view (Figure \@ref(fig:hierarchical)a). We see 
in this case that is difficult
to see the sub-clusters in the tour view, however, once we zoom and highlight
they become more apparent (Figure \@ref(fig:hierarchical)b). When we brush 
the sub-clusters in the t-SNE, their
relative placement is again exaggerated, with the tour showing that they
are indeed much closer than the impression the t-SNE view gives.

@@

Next we explore some simulated noisy tree structured data (Figure
\@ref(fig:fake-trees)). Our interest here is how t-SNE visualisations break
topology of the data, and then seeing if we can resolve this by tweaking the
default parameters with reference to the global view of the data set. 
This simulation aims to mimic branching trajectories of cell differentiation:
if there were only mature cells, we would just see the tips of the branches 
which have a hierarchical pattern of clustering.

@@

Moreover, we run t-SNE using the default arguments on the complete data (this
keeps the first 50 PCs, sets the perplexity to equal 30 and performs random
initialisation). We then create a linked
tour with t-SNE layout with **liminal** as shown in Figure \@ref(fig:trees-01).

@@

From the linked video, we see that the t-SNE view has been unable to 
recapitulate the topology of the tree - the backbone (blue) branch has been 
split into three fragments (Figure \@ref(fig:trees-01)a). We can see 
this immediately via the linked highlighting over both
plots. If we click on the legend for the zero branch, the blue coloured points
on each view are highlighted and the remaining points are made transparent.
From, here it becomes apparent from the tour view that the blue branch forms
the backbone of the tree and is connected to all other branches. From the video 
it is easy to see that cluster sizes formed via t-SNE can be misleading; from the tour
view there is a lot of noise along the branches, while this does not appear to
be the case for the t-SNE result (Figure \@ref(fig:trees-01)b).

From the first view, we modify the inputs to the t-SNE view, to try and produce
a better trade-off between local structure and retain the topology of the data.
We keep every parameter the same except that we initialise $Y$ with the first
two PCs (scaled to have standard deviation 1e-4) instead of the default random
initialisation and increase the perplexity from 30 to 100. We then combine
these results with our tour view as displayed in the linked video in the caption
of Figure \@ref(fig:trees-02).

@@

The video linked in Figure \@ref(fig:trees-02) shows that this selection of 
parameters results in the 
tips of the branches (the three black dots in Figure \@ref(fig:fake-trees)a) 
being split into three clusters representing the terminal branches of the tree. 
However, there are perceptual issues following the placement of the three 
groupings on the t-SNE view that become apparent via simple linked brushing. 
If we brush the
tips of the yellow and brown branches (which appear to be close to each other
on the t-SNE view), we immediately see the placement is distorted in the t-SNE
view, and in the tour view these tips are at opposite ends of the tree 
(Figure \@ref(fig:trees-02)b).
Although, this is a known issue of the t-SNE algorithm, we can easily identify
it via simple interactivity without knowing the inner workings of the method.

@@

Due to latency of the **liminal** interface we do a weighted sample of  the rows based on
cluster membership, leaving us with approximately 10 per cent of the original
data size - 4,590 cells. Although this is not ideal, it still allows us to
get a sense of the shape of the clusters as seen from the linked video in Figure \@ref(fig:mouse-01). If one was interested in performing more in-depth cluster analysis we recommend an iterative approach of removing large clusters and then re-running the **liminal** view as a way finding more granular cluster structure.

@@

From the video linked in Figure \@ref(fig:mouse-01), we learn that the embedding has mostly captured the
clusters relative location to each other to their location in high dimensional space, with a notable exception of points in cluster 3 and 10 as shown with
linked brushing (Figure \@ref(fig:mouse-01)a). As expected, t-SNE mitigates the crowding problem that is
an issue for tour in this case, where points in clusters 2,4,6, and 11 are
clumped together in tour view, but are blown up in the embedding view (Figure \@ref(fig:mouse-01)b). 
The tour appears to form a tetrahedron like shape, with points lying on
surface and along the vertices of the tetrahedron in 5-$d$ PCA space -
a phenomena that has also been observed in @Korem2015-af 
(Figure \@ref(fig:mouse-01)c). Brushing on
the tour view, reveals points in cluster 9 that are diffuse in the embedding 
view, points in cluster 9 are relatively far away and spread apart from other 
clusters in the tour view, but has points placed in cluster 3 and 9 in the embedding
(Figure \@ref(fig:mouse-01)d).

@@

From the video linked in Figure \@ref(fig:mouse-02), we see that the 
expression of the marker genes, appear to separate
the previously clumped clusters 2,4,6, and 11 from the other clusters, 
indicating that these genes are expressed in all four clusters (Figure \@ref(fig:mouse-02)a). After zooming, we can see a trajectory forming along
the clusters, while the axis view shows that magnitude of expression in 
the marker genes is similar across these separated clusters which is 
consistent with the results of marker gene analysis (Figure \@ref(fig:mouse-02)b).
